---
title: "sigvar analysis of samples with varying carcinogen exposure"
author: "Maike Morrison"
date: "`r Sys.Date()`"
# output: rmarkdown::html_vignette
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{carcinogen_figure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.width = 10
)
```




# Introduction

This vignette reproduces the analysis conducted in Figure 2 of [Morrison et al.](https://doi.org/10.1101/2023.11.23.23298821). 

```{r setup}
library(tidyverse)
library(patchwork)
library(sigvar)
library(ggtext)
pal =   c("#264653", "#2a9d8f", "#e9c46a", "#f4a261", "#e76f51") %>%
  `names<-`(c("Non-smoker", "Unexposed to\nasbestos", "Passive smoker", "Exposed to\nasbestos", "Smoker"))
```


# Rename and format data

## Riva et al. mice carcinogen data

```{r}
riva_sig_defs = sigvar::mutsig_carcinogens_mice_SBS.refs

chem_abbreviations = data.frame(chemical = c("1-trans-delta-9-TetrahydrocannabJnol", "1,2,3-Trichloropropane", "antimony trioxide", "antrhaquinone", "Black Cohosh", "bromochloroacetic acid", "cobalt metal", "CUMENE", "DIETHANOLAMINE", "furan", "ginkgo biloba extract", "Indium phosphide", "isobutyl nitrite", "nickel oxide", "nickel subsulfite", "nickel sulfate hexahydrate", "OXAZEPAM", "p-Nitroaniline", "p Nitrobenzoic acid'", "Pentabromodiphenyl Ether Mixture", "primaclone", "Resveratrol", "Salicylazosulfapyridine", "Scopolamine hydrobromide trihydrate", "sodium tungstate dihydrate", "Sodium Tungstate Dihydrate", "Triethanolamine", "vanadium pentoxide", "vinylidene chloride"),
                                chem_name = c("1-trans-delta-9-TetrahydrocannabJnol", "TCP", "Antimony trioxide", "Anthraquinone", "Black cohosh", "Bromochloroacetic\nacid", "Cobalt metal", "Cumene", "Diethanolamine", "Furan", "G. biloba extract", "Indium phosphide", "Isobutyl nitrite", "Nickel oxide", "Nickel subsulfide", "Nickel sulfate\nhexahydrate", "Oxazepam", "p-Nitroaniline", "p Nitrobenzoic acid'", "DE-71", "Primaclone", "Resveratrol", "Salicylazosulfapyridine", "Scopolamine hydrobromide trihydrate", "Sodium tungstate\ndihydrate", "Sodium tungstate\ndihydrate", "Triethanolamine", "Vanadium pentoxide", "VDC"))


riva_relab = mutsig_carcinogens_mice_SBS %>% left_join(chem_abbreviations) %>%
  select(Sample, chemical = chem_name, dose_numeric, Tissue, mSBS1:mSBS_N3) %>%
  filter(Tissue %in% c("liver", "lung")) 


riva_relab$chemical[stringr::str_detect(riva_relab$Sample, "SPONTANEOUS")] = "Spontaneous"
chems = c("Spontaneous", setdiff( riva_relab$chemical %>% unique, "Spontaneous"))

riva_relab$chemical = factor(riva_relab$chemical, ordered = TRUE, levels = chems)

# filter to carcinogen X tumor combinations with more than 5 samples
riva_relab_counts = riva_relab %>% group_by(Tissue, chemical) %>%
  summarise(n = n()) %>% filter(n>1) 

sigs = colnames(riva_relab)[5:15]

riva_cossim <- lsa::cosine(x = as.matrix(riva_sig_defs[,-c(1:2)]))[sigs, sigs]

riva_var = sigvar(sig_activity = riva_relab, 
                   K = 11, S = riva_cossim,
                   group = c("chemical", "Tissue")) %>% 
  `colnames<-`(c("chemical", "Tissue", "group", "across", "mean_within"))

riva_within = riva_relab %>% select(chemical, Tissue) %>%
  mutate(within = apply(riva_relab, 1, het, K = 11, S = riva_cossim)) %>%
  left_join(riva_var)

```



## Zhang et al. never-smokers data

```{r PROCESS DATA FROM ZHANG ET AL (Never smokers)}
zhang = zhang_sig_activity

zhang_sig_cossim = zhang_sig_similarity
```

## Chen et al. East-Asian lung data

```{r PROCESS DATA FROM CHEN ET AL (EAST ASIANS)}
chen = smoker_sigs_chen

chen_sig_cossim = smoker_sigs_chen_cossim
```

## MESOMICS data

```{r}
meso_raw = MESOMICS_CN_SBS[,str_detect(colnames(MESOMICS_CN_SBS),"^CN[0-9]+") ]
meso_raw = sweep(meso_raw,1,rowSums(meso_raw),"/")

meso = bind_cols(
  Asbestos_exposure=MESOMICS_CN_SBS$Professional.Asbestos,  
  as_tibble(meso_raw))  %>%
  filter((Asbestos_exposure %in% c("Exposed", "Non exposed"))) %>% 
  na.omit() %>%
  mutate(Asbestos_exposure_formatted = ifelse(Asbestos_exposure == "Exposed", 
                                              "Exposed to\nasbestos", 
                                              "Unexposed to\nasbestos"),
         .before = 2)


meso_cossim <- lsa::cosine(x = COSMIC3.1_CN[,-1] %>% as.matrix)

```



# Statistics
```{r}
zhang_label = "**Lung cancer in<br>never-smokers**<br>(Zhang et al. 2021)"
chen_label = "**Lung<br>adenocarcinoma<br>in East Asians**<br>(Chen et al. 2020)"
meso_label = "**Mesothelioma**<br>(Mangiante et al. 2023)"
carcinogens_within = rbind(data.frame(group = zhang$passive_smoking, 
                 within = apply(zhang, 1, het, K = 14, S = zhang_sig_cossim),
                 data = zhang_label),
      
      data.frame(group = chen$Smoker, 
                 within = apply(chen, 1, het, K = 3, S = chen_sig_cossim),
                 data = chen_label),
      
      data.frame(group = meso$Asbestos_exposure_formatted, 
                 within = apply(meso, 1, het, K = 7, S = meso_cossim),
                 data = meso_label))

zhang_var = sigvar(sig_activity = zhang, K = 14,
                   group = "passive_smoking", S = zhang_sig_cossim) %>% 
  `colnames<-`(c("group", "across", "mean_within")) %>%
  mutate(data = zhang_label)
chen_var = sigvar(sig_activity = chen, K = 3, group = "Smoker",
                  S = chen_sig_cossim) %>%
  `colnames<-`(c("group", "across", "mean_within")) %>%
  mutate(data = chen_label)
meso_var = sigvar(sig_activity = meso %>% na.omit, K = 7, 
                  group = "Asbestos_exposure_formatted", S = meso_cossim) %>% 
  `colnames<-`(c("group", "across", "mean_within")) %>%
  mutate(data = meso_label)

carcinogens = rbind(zhang_var, chen_var, meso_var)

carcinogens_all = left_join(carcinogens_within, carcinogens)

```


# Plot
```{r}
group_labels = carcinogens
nudge = 0.025
group_labels$across = group_labels$across + 
  c(-nudge, nudge, nudge*0.8, - nudge*0.8, nudge*1.4, -nudge*1.4)
group_labels = group_labels %>% mutate(max_mean_within= max(mean_within)+.13)

data_labels = carcinogens %>% group_by(data) %>%
  summarise(across = mean(across), mean_within = mean(mean_within))

data_labels$across = data_labels$across + c(0, 0.01, 0)

carcinogens_plot =
  ggplot(carcinogens_all) + 
    
  geom_segment(aes(y = across, yend = across, color = group,
                   x = min, xend = max), 
               size = 1, alpha = 0.4, 
               data = carcinogens_all %>% group_by(paste0(data, group)) %>% 
                 summarise(min = min(within), max = max(within),
                           across, data, group) %>% 
                 distinct) +

    geom_richtext(aes(x = -0.18, y = across, label = data), data = data_labels,
                  hjust = 0, 
                  fill = NA, label.color = NA, # remove background and outline
                  size = 3) +
  
   geom_violin(aes(x = within, y = across, fill = group,
                   group = paste0(group, data)), color = NA, 
               alpha = 0.4, width = 0.05) +
  
  geom_line(aes(x = mean_within, y = across, group = data), size = 1,
            color = "#7E7E7E", alpha = 0.5) + 
  geom_point(aes(x = mean_within, y = across, color = group), size = 3)+
  
  theme_bw() + 
  theme(panel.grid = element_blank(),
        legend.position = "none") +
  
  geom_text(aes(x = max_mean_within, y = across, label = group, color = group),
            data = group_labels, 
            fontface = "bold", nudge_x = 0.16, size = 3, hjust = 1) +

  scale_color_manual(values = pal) +
  scale_fill_manual(values = pal) +
  scale_x_continuous(breaks = c(0, 0.2, 0.4, 0.6)) +
  xlab("Within-sample diversity") +
  ylab("Across-sample heterogeneity") + coord_cartesian(xlim = c(-0.145, 0.63))
carcinogens_plot
```



## Significance - bootstrapping

The below step is computationally expensive, so we load the result as a file included in the sigvar R package. 

```{r}
# tictoc::tic()
# # Riva stats
# riva_bootstrap_list_greater = list()
# chemical_tissue = list()
# i = 1
# set.seed(1)
# for(chem in setdiff(chems, "Spontaneous")){
#   riva_chem_spontaneous = filter(riva_relab, chemical %in% c("Spontaneous", chem))
#   riva_chem = filter(riva_relab, chemical == chem)
#   for(tissue in unique(riva_chem$Tissue)){
#     riva_bootstrap_list_greater[[i]] = sigvar::sigboot(sig_activity = riva_chem_spontaneous %>% filter(Tissue == tissue),
#                    K = 11, S = riva_cossim,
#                    group = c("chemical", "Tissue"), n_replicates = 1000, seed = 1, alternative = "greater")
#     chemical_tissue[[i]] = c()
#     i = i + 1
#   }
# }
# tictoc::toc()
# beepr::beep()
# saveRDS(riva_bootstrap_list_greater, file = "sim_out/mice_boot_stats_seed1_greater.rds")
# readRDS("sim_out/mice_boot_stats_seed1_greater.rds")

# chemical_tissue = sapply(riva_bootstrap_list_greater, 
#                          function(list) list$observed_stats$group[!stringr::str_detect(list$observed_stats$group,
#                                                                                        "Spontaneous")])
# 
# group_table = do.call(rbind, lapply(riva_bootstrap_list_greater, function(list) list$observed_stats[,1:3])) %>% 
#   distinct
# 
# # riva_p_values_less = right_join(group_table, 
# #                            do.call(rbind, lapply(riva_bootstrap_list_less, function(list) list$P_value)) %>% data.frame %>%
# #   mutate(group = chemical_tissue, .before = 1))
# 
# riva_p_values_greater_raw = do.call(rbind, lapply(riva_bootstrap_list_greater, function(list) list$P_value))
# 
# riva_p_values_greater = rbind(riva_p_values_greater_raw %>% data.frame %>% filter(group_2 %in%  c("Spontaneous_lung", "Spontaneous_liver")),
#   (riva_p_values_greater_raw %>% data.frame %>% filter(group_1 %in%  c("Spontaneous_lung", "Spontaneous_liver")) %>%
#     mutate(across(3:5, function(x) 1 - as.numeric(x))))[,c(2,1,3:5)] %>% `colnames<-`(colnames(riva_p_values_greater_raw)))
# 
# riva_p_values_greater[riva_p_values_greater == "<0.001"] = 0
# 
# riva_p_values_greater[,3:5] = apply(riva_p_values_greater[,3:5], c(1,2), as.numeric)
# 
# riva_p_values_greater = riva_p_values_greater %>% right_join(group_table, ., by = c("group" = "group_1"))

# saveRDS(riva_p_values_greater, "sim_out/mice_boot_stats_seed1_greater_clean.rds")

# ^ This file is saved as "mutsig_carcinogens_mice_bootstrap_p_vals" in sigvar

```

```{r}
signif_pal = c("#DABD61FF", "#D95F30FF", "#BE3428FF", "#620800") %>% 
    `names<-`(c("\U00B7", "*", "**", "***"))

sig2 = mutsig_carcinogens_mice_bootstrap_p_vals
sig2$chemical = as.character(sig2$chemical)

sig2$significance = 
  ifelse(sig2$mean_within_sample_diversity<.001, "***",
         ifelse(sig2$mean_within_sample_diversity<.01, "**", 
                ifelse(sig2$mean_within_sample_diversity<.05, "*",
                       ifelse(sig2$mean_within_sample_diversity<0.1, "\U00B7",
                              ""))))


sig2$within_sig = ifelse(sig2$Tissue == "liver", 0.45, 0.5)

bold_list = sig2 %>% group_by(chemical) %>% 
  summarise(min_within = min(mean_within_sample_diversity)) %>% 
  mutate(chemical_bold = ifelse(min_within < 0.1, paste0("**", chemical, "**"),
                                as.character(chemical))) %>% 
  select(chemical, chemical_bold) %>% distinct() %>%
  filter(chemical != "Spontaneous") %>%
  rbind(c("Spontaneous", "**Spontaneous**"))


riva_within_sig2  = left_join(riva_within, sig2, 
                              by = c("chemical", "Tissue", "group")) %>%
  left_join(bold_list)

sig2 = left_join(sig2, bold_list)


chems_ordered = riva_within_sig2 %>% filter(Tissue == "liver") %>%
  arrange(-mean_within) %>%
  select(chemical_bold) %>% unique %>% unlist

riva_within_sig2$chemical_ordered = factor(riva_within_sig2$chemical_bold, 
                                           ordered = TRUE,
                                           levels = c(setdiff(chems_ordered,
                                                              "**Spontaneous**"), 
                                                      "**Spontaneous**"))

```



```{r}
riva_1d = ggplot(riva_within_sig2, aes(y = chemical_ordered, x = within)) +
  
    geom_point(aes(x = mean_within, y = chemical_ordered), size = 3, alpha = 0) +

  geom_vline(aes(xintercept = mean_within),
             data = riva_within_sig2 %>% filter(chemical == "Spontaneous"), 
             linewidth = 1, color = "#3A488AFF", alpha = 0.7) +
   
  geom_violin(fill = "#3A488AFF", color = NA, width = 0.9, alpha = 0.7,
             data = riva_within_sig2 %>% filter(chemical == "Spontaneous")) + 
  geom_violin(aes(fill = significance), color = NA, width = 1.3, alpha = 0.7,
             data = riva_within_sig2 %>% filter(chemical != "Spontaneous")) + 
  geom_point(aes(x = mean_within, y = chemical_bold), size = 2.5, alpha = 0.3) +
  
  facet_grid(~ifelse(Tissue == "liver", 
                     "**Mouse liver tumors**<br>(Riva et al. 2020)", 
                     "**Mouse lung tumors**<br>(Riva et al. 2020)"), 
             scales = "free_x", space = "free_x") + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        strip.background = element_blank(),
        strip.text = element_markdown(size = 9),
        legend.position = "none",
        axis.title.y = element_blank(),
        axis.text.y = element_markdown()) +
  xlab("Within-sample diversity") +
  
  geom_text(aes(x = within_sig, y = chemical_bold, 
                label = significance, color = significance), 
            data = riva_within_sig2, size = 5,
            fontface = "bold", vjust = 0.5, hjust = 1) +
  scale_fill_manual(values = signif_pal) +
  scale_color_manual(values = signif_pal) + 
  
  geom_richtext(
    data = data.frame(label = "
  <span style='color:#DABD61FF'> \U00B7   <i>P</i> < 0.10</span><br>
  <span style='color:#D95F30FF'>*  <i>P</i> < 0.05</span><br>
  <span style='color:#BE3428FF'>**    <i>P</i> < 0.01</span><br>
   <span style='color:#620800'>***    <i>P</i> < 0.001</span>",
  Tissue = "liver"), 
            aes(x = 0.34, y = "Nickel oxide", label = label), color = "grey")


``` 



# Add arrows
```{r, fig.width = 20, fig.height = 8}
within_center = 0.6025
across_center = 0.547

arrows_df <- data.frame(xstart = c(0.08, 0.08, within_center - 0.2,
                                   within_center + 0.2), 
                        xend   = c(0.08, 0.08, within_center - 0.42, 
                                   within_center + 0.42), 
                        ystart = c(across_center - 0.2, across_center + 0.2, 
                                   0.018, 0.018), 
                        yend   = c(across_center - 0.42, across_center + 0.42,
                                   0.018, 0.018), 
                        xlab   = c(-0.01, -0.01, within_center - 0.31, 
                                   within_center + 0.31),
                        ylab   = c(across_center - 0.31, across_center + 0.31, 
                                   -0.025, -0.025),
                        color  = c("blue", "orange", "blue", "orange"),
                        label  = c("Identical\ntumors ",
                                  "Dissimilar\ntumors ", 
                                  "Few signatures", 
                                  "Many signatures"))
base <- ggplot(arrows_df) + 
  geom_segment(aes(x = xstart, y = ystart, xend = xend,
                   yend = yend, color = color),
            arrow = arrow(length = unit(0.3, "cm")), 
            lwd=1.2, color = "#B3B3B3") +
  geom_text(aes(x = xlab, y = ylab , label = label, color = color),
            lineheight = 0.9, color = "#676767") +
  theme_void() + 
  theme(legend.position = "none", 
        plot.background = element_rect(fill = "white", color = "white")) + 
  coord_cartesian(xlim = c(-0.05,1), ylim = c(0,1.02))



# CARCINOGENS
carcinogens_arrows <- base +
  annotation_custom(grob = ggplotGrob(carcinogens_plot + 
                                        xlab("Mean within-sample\ndiversity") + 
                                        ylab("Across-sample\nheterogeneity") + 
                                        theme(#axis.title = element_blank(),
                                          plot.background = element_blank())), 
                    xmin = 0, xmax = 1.05, ymin = -0.09, ymax = 1.03)
carcinogens_arrows



riva_center = 0.619

base_within_only <- ggplot(data.frame(xstart = riva_center + c(-0.15, 0.15), 
                        xend   = riva_center + c(-0.35, 0.35), 
                        ystart = c(0.031, 0.031), 
                        yend   = c(0.031, 0.031), 
                        xlab   = riva_center + c(-0.25, 0.25),
                        ylab   = c(-0.023, -0.023),
                        color  = c("blue", "orange"),
                        label  = c("Few signatures", 
                                  "Many signatures"))) + 
  theme_void() + 
  theme(legend.position = "none",
        plot.background = element_rect(fill = "white", color = "white")) + 
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))



# RIVA

riva_arrows <- base_within_only +
  annotation_custom(grob = ggplotGrob(riva_1d+ 
                                        xlab("Mean within-sample\ndiversity") + 
                                        ylab("Across-sample\nheterogeneity") + 
                                        theme(#axis.title = element_blank(),
                                          plot.background = element_blank())), 
                    xmin = -0.09, xmax = 1, ymin = -0.095, ymax = 1.13)
# riva_arrows



panels = patchwork::wrap_plots((riva_arrows), (carcinogens_arrows), 
                               widths = c(1.3, 1)) + 
  plot_annotation(tag_levels = "a")

panels

```




