---
title: "sigFAVA analysis of the mutational signatures of carcinogens"
author:
  - name: "Nicolas Alcala"
    affiliation: "Rare Cancers Genomics Team, International Agency for Research on Cancer / World Health Organization"
    email: "alcalan@iarc.who.int"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Mutational signatures of carcinogens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sigFAVA)
```

# Introduction

This vignette details an example of using sigFAVA to detect the effect of exposure to a carcinogen on the variability of mutational signatures. 

# Datasets

## Mutational signatures of tumors from mice exposed to known or suspected carcinogens (Riva et al. Nat Genet 2020)

```{r }

```

## Mutational signatures of malignant pleural mesothelioma patients with various degrees of asbestos exposure (Mangiante et al. Nat Genet 2023)

```{r }

```

# Computing variability statistics

## Gini-Simpson index
```{r }

```

## FST

# Plotting
```{r }

```

# Session information

```{r }
sessionInfo()
```

